---
title: "Tutorial: Resources in R"
author:
- name: Yannick Marcon
  affiliation:
  - &epigeny Epigeny
  email: yannick.marcon@obiba.org
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Tutorial: Resources in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
BiocStyle::markdown()
options(width=100)
knitr::opts_chunk$set(comment="", warning=FALSE, message=FALSE, cache=TRUE)
```

# The resourcer Package

The [resourcer](https://cran.r-project.org/package=resourcer) package is an R implementation of the concept of Resources (see also the <a href="https://github.com/obiba/resourcer" target="_blank">source repository</a>).

## Installation

Install the package if not already available in the R environment.

```{r install, message=FALSE}
if (!require(resourcer)) {
  install.packages(c("resourcer"))
}
```

## Load

When loading the [resourcer](https://cran.r-project.org/package=resourcer) package, the output informs about the available resource **resolvers**. These are responsible for building the appropriate connector to a resource object. This set of resolvers can be extended to new types of resources.

```{r load, message=FALSE}
library(resourcer)
```

Note that the `resourcer` package makes an extensive use of the [R6](https://cran.r-project.org/package=R6) class system.

# Building a Data Resource Object

The resource gives access to some data.

## CSV File Resource

The file is publicly available on the Github website: [preview of CAPostalCodes.csv](https://github.com/obiba/obiba-home/blob/master/opal/seed/fs/home/administrator/geo/CAPostalCodes.csv). 

```{r CAPostalCodes.res}
CAPostalCodes.res <- resourcer::newResource(
  url = "https://github.com/obiba/obiba-home/raw/master/opal/seed/fs/home/administrator/geo/CAPostalCodes.csv", 
  format = "csv"
)
CAPostalCodes.res
```

Note that at this point, no connection, nor data extraction has been performed. It is a simple definition object.

The following code will resolve the resource and build the corresponding client connector.

```{r CAPostalCodes.client}
CAPostalCodes.client <- resourcer::newResourceClient(CAPostalCodes.res)
class(CAPostalCodes.client)
```

The resource was identified as a "tidy" data file, i.e. data that can be read using one of the reader developed by the [tidyverse](https://www.tidyverse.org/) project. In the case of the `csv` data format, the [readr](https://readr.tidyverse.org/) package is used. This CSV reader tries to guess the data type of the columns.

At this point again, no connection with the remote server has been formally established.

This client function call gives access to the data as a `data.frame`:

```{r CAPostalCodes.data, message=FALSE}
CAPostalCodes.data <- CAPostalCodes.client$asDataFrame()
head(CAPostalCodes.data)
```

It is also possible to coerce a resource object directly to a `data.frame`, without explicitly building a resource client object. It is as simple as:

```{r CAPostalCodes.data2, message=FALSE}
CAPostalCodes.data <- as.data.frame(CAPostalCodes.res)
head(CAPostalCodes.data)
```


## R Data File Resource

The file is stored in a Opal server file system. Authentication and authorization apply and is performed with a [Personal Access Token](http://opaldoc.obiba.org/en/latest/web-user-guide/my-profile.html#personal-access-tokens).

```{r gps_participant.res}
gps_participant.res <- resourcer::newResource(
  url = "opal+https://opal-demo.obiba.org/ws/files/projects/RSRC/gps_participant.RData",
  format = "data.frame",
  secret = "EeTtQGIob6haio5bx6FUfVvIGkeZJfGq"
)
gps_participant.res
```
Make a resource client object.

```{r gps_participant.client}
gps_participant.client <- resourcer::newResourceClient(gps_participant.res)
class(gps_participant.client)
```

The resource was identified as an R data file, containing a `data.frame` object.

When extracting the inner R object from the remote R data file, the resource client object will establish the connection with the Opal server and will authenticate with the provided Personal Access Token; then it will perform the file download and will read its content. The function ``getValue()`` returns the raw object (in this case a `data.frame`)

```{r gps_participant.data}
gps_participant.data <- gps_participant.client$getValue()
head(gps_participant.data)
```

There are no limitations regarding the class of the object contained in the R data file. The only requirements are the ones of the `base::load()` function, i.e. the library in which the class of the object is defined must be available in the R environment.

# Building a Computation Resource Object

The resource gives access to some remote computation services.

## SSH Resource

A server is accessible through a secure shell. The path part of the URL is the remote working directory. The available commands are defined by the `exec` query parameter.

```{r ssh.res}
ssh.res <- resourcer::newResource(
  url = "ssh://plink-demo.obiba.org:2222/home/master/brge?exec=ls,pwd",
  identity = "master",
  secret = "master"
)
```

The resource connection client is resolved as follow:

```{r ssh.client}
ssh.client <- resourcer::newResourceClient(ssh.res)
class(ssh.client)
```

This type of client allows to issue shell commands through a SSH connection.

```{r ssh.cmds}
ssh.client$getAllowedCommands()
```

Trying to coerce to a `data.frame` raises an error, because there is no tabular data representation of such a resource:

```{r ssh.df}
tryCatch(ssh.client$asDataFrame(), error = function(e) e)
```

To execute a remote shell command:

```{r ssh.cmd, message=FALSE, warning=FALSE, error=FALSE}
rval <- ssh.client$exec("ls", "-la")
rval
```

The resulting value contains different information:

* `status` of the command (failure if not 0),
* `output` the character vector of the command output,
* `error` is the error message if command failed,
* `command` is the actual shell command that was executed.

For example some bad shell command arguments would return a value with an error:

```{r ssh.cmd2, message=FALSE}
rval <- ssh.client$exec("ls", "-xyz")
rval
```

Calling a shell command that is not allowed would raise an error.

```{r ssh.cmd3, message=FALSE}
tryCatch(ssh.client$exec("plink"), error = function(e) e)
```

# More About Resources

## Extending Resources

The `resourcer` package comes with some built-in resource types. These can be extended by programming your own resource resolver and client. For more information, read the sections about Resources in the book [Orchestrating privacy-protected non-disclosive big data analyses of data from different resources with R and DataSHIELD](https://isglobal-brge.github.io/resource_bookdown).

## Resources in Opal and DataSHIELD

The Opal data management server facilitates the usage of resources by:

* providing a graphical user interface for declaring resources (defining the resource's URL can be tricky),
* discovering which type of resources are available in the R server environment,
* granting permissions to use the resource,
* assigning resources in a R server session,
* integrating resources to DataSHIELD.

See also <a href="https://rpubs.com/jrgonzalezISGlobal/tutorial_resources" target="_blank">Tutorial: Using the resources in Opal and DataSHIELD</a>.
